### 2021-08-22 plos_latex_template.tex
RJ edited organizing team and search for people rules

This article suggests rules to pivot traditional conferences towards inclusiveness and diversity and to welcome minoritized groups. These tips stem from the authors' experience of organizing a virtual and global conference for users and developers of the R programming language, useR! 2021. We embraced the challenge of organizing a high-quality conference in the context of the COVID-19 pandemic using the chance of a virtual conference environment to create a kind, truly inclusive, accessible, and welcoming experience for everyone. Most of the authors also have experience of  organizing other regional and national academic conferences and events in communities of practice such as R-Ladies and The Carpentries. Here, we share the lessons learned within the past year of organizing this global useR! conference, summarized as 10 simple rules towards an inclusive conference.

@@

As a starting point, identify which groups are usually unsupported or discriminated against in your community or event, and determine the contributing factors. Next, think about set measurable indicators of success when organizing a diverse conference. The numbers resulting from these indicators are not the primary goal. Instead, this vision should guide and help hold the organizing team accountable towards these diverse and inclusive goals.

@@

In-person interaction is priceless; however, it is more expensive for some, even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arend_disparity_2019, gewin_what_2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipour_evaluating_2020, niner_better_2021}; in a study case, conference attendance accounted for $35\%$ of a researchers footprint .

@@

\section{Rule X: Have an inclusive and diverse organizing team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually underrepresented groups (see \textbf{Rule 1} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding of the challenges in different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. Even if creating and maintaining such an interdisciplinary team is challenging, the positive outcomes far outweigh any minor inconveniences, and the team will grow, learn, and embrace more inclusive practices. For instance, if the organizing team is global and remote, time zones can be hard to manage, with meetings scheduled at odd hours. The team may also have to review carefully and teach themselves the correct vocabulary for internal and external communication and identify the best ways to account for every culture and situation. 
An inclusive team will let everyone know if the organization is failing in any aspect of inclusiveness or accessibility, and help improve it \cite{hong_groups_2004}.

@@

\section{Rule X: Search proactively the best people for each task}

While this rule may seem intuitive, one would be surprised by how often conference organizers tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in. For example, a cisgender or abled person estimating what might work best for transgender people or people with sensory disabilities. 
Therefore, to build your team of organizers for the conference, you need to actively seek out and recruit the best people for each task. The intention to put together a diverse team alone is not enough. One needs to look for, reach out to, invite, encourage, and onboard these people until there is ample representation across the diversity spectrum and dimensions [\textbf{Fig. 1}]. To achieve this, it will be necessary to go beyond one's limited networks when inviting people to the organizing team, keynote speakers, program committee, session chairs, reviewers, and other roles. Ensuring diversity in each of these teams and panels is a deliberate process -- if there is not much diversity in the first set of names, it needs to be revised to include people from minoritized groups. If these groups aren't showing up in the A-list, it is seldom because of their technical acumen -- it is because we, as organizers, are not used to turning every work and looking in unfamiliar places to bring in the talent.  Many of our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyer_notice_2021, sarabipour_evaluating_2020}, thus encouraging us to look further to find great people that are not routinely in the spotlight. We need to go beyond our narrow and often limited networks. So, as long as we start with an inclusive and diverse organizing team (\textbf{Rule 2}), we can aim at finding the right people for various aspects of the conference organization. The regional and local communities are also good sources to tap into. For example, for useR! 2021, groups like AfricaR, R-Ladies, MiR, Forwards, and LatinR were fundamental to reach people for the organizing team, potential attendees, and sponsors.

Disabled people often repeat "Nothing about us without us". This means the actual life experiences of people from URM are not replaceable with good intentions from people outside these groups. The same holds for other dimensions of diversity. No space will be truly inclusive and welcoming to URM without people from these groups bringing their experience to bear. This means people in privileged groups may need to step down from positions of power. Of course, no-one has the same experiences and experiences the same oppressions in life. %When building a diverse team, bear in mind that

@@

Conferences are often inaccessible to people with hearing, visual, and other disabilities \cite{price_access_2009}. People may also find conferences inaccessible due to their caring responsibilities, dietary restrictions, religious practices, or LGBTQ+ status. Each of these groups should be considered at every stage of the planning process. Welcoming representatives from these groups into the organizing team would allow them to take part in the decisions from the beginning and spot right away the inaccessible practices that need to improve (see \textbf{Rule 2}). If the conference is in person, the venue must be accessible for people with mobility limitations. Accommodations such as a quiet space for neurodivergent people and a space for breastfeeding should also be provided. Images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. Platforms for conference registration and abstract submission, websites and chat platforms should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). These aspects should be tested well in advance of going live. Captioning for presentations should be available in more than one language if possible. If the conference is in-person, presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. Accessibility guidelines for slides and presentations should be provided to the presenters, and their use should be encouraged (see \url{https://user2021.r-project.org/participation/accessibility/} for example). Among other things, the speakers should provide raw and accessible material to their talks before the conference, e.g., R Markdown, HTML, or \TeX{} files; if they prerecord their presentation, they should include their video and ensure that their face is visible so that deaf and hard-of-hearing people can read their lips if needed. Adding captions to recorded videos is relatively easy these days and is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general.  Accessibility awards at the end of the conference are an excellent way to acknowledge the speakers who were mindful of inclusiveness when preparing and delivering their talks.

>>>>

This article suggests rules to pivot traditional conferences towards inclusiveness and diversity and to welcome minoritized groups. These tips stem from the authors' experience of organizing useR! 2021, a virtual and global conference for users and developers of the R programming language \cite{r_core_team_2021}. We embraced the challenge of organizing a high-quality conference in the context of the COVID-19 pandemic using the chance of a virtual conference environment to create a kind, truly inclusive, accessible, and welcoming experience for everyone. Most of the authors also have experience of  organizing other regional and national academic conferences and events in communities of practice such as R-Ladies and The Carpentries. Here, we share the lessons learned within the past year of organizing this global useR! conference, summarized as 10 simple rules towards an inclusive conference.

@@

As a starting point, identify which groups are usually unsupported or discriminated against in your community or event, and determine the contributing factors.

Next, think about set measurable indicators of success when organizing a diverse conference. The numbers resulting from these indicators are not the primary goal. Instead, this vision should guide and help hold the organizing team accountable towards these diverse and inclusive goals.

@@

In-person interaction is priceless; however, it is more expensive for some, even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arend_disparity_2019, gewin_what_2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinson_r_medicine_2021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipour_evaluating_2020, niner_better_2021}; in a study case, conference attendance accounted for $35\%$ of a researchers footprint .

@@

\section{Rule X: Have an inclusive and diverse organizing team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually underrepresented groups (see \textbf{Rule 1} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding of the challenges in different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. But this will only work if there is real inclusion. One would be surprised by how often conference leads tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in. For example, a cisgender or abled person estimating what might work best for transgender people or people with sensory disabilities. The same holds for other dimensions of diversity. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. Even if creating and maintaining such a team and space is challenging, the positive outcomes far outweigh any minor inconveniences (e.g. time zones), and the team will grow, learn, and embrace more inclusive practices \cite{hong_groups_2004}.

%The team may also have to review carefully and teach themselves the correct vocabulary for internal and external communication and identify the best ways to account for every culture and situation. % rj: left this out, Heather said it doesn't make much sense here

@@

Conferences are often inaccessible to people with hearing, visual, and other disabilities \cite{price_access_2009}. People may also find conferences inaccessible due to their caring responsibilities, dietary restrictions, religious practices, or LGBTQ+ status. Each of these groups should be considered at every stage of the planning process. Welcoming people from these groups into the organizing team would allow them to take part in the decisions from the beginning and spot right away the inaccessible practices that need to improve (see \textbf{Rule 2}). If the conference is in person, the venue must be accessible for people with mobility limitations. Accommodations such as a quiet space for neurodivergent people and a space for breastfeeding should also be provided. Images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. Platforms for conference registration and abstract submission, websites and chat platforms should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). These aspects should be tested well in advance of going live. Captioning for presentations should be available in more than one language if possible. If the conference is in-person, presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. Accessibility guidelines for slides and presentations should be provided to the presenters, and their use should be encouraged (see \url{https://user2021.r-project.org/participation/accessibility/} for example). Among other things, the speakers should provide raw and accessible material to their talks before the conference, e.g., R Markdown, HTML, or \TeX{} files; if they prerecord their presentation, they should include their video and ensure that their face is visible so that deaf and hard-of-hearing people can read their lips if needed. Adding captions to recorded videos is relatively easy these days and is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general.  Accessibility awards at the end of the conference are an excellent way to acknowledge the speakers who were mindful of inclusiveness when preparing and delivering their talks.

\section{Rule X: Search proactively the best people for invited roles}

When inviting people as keynote speakers, program committee, session chairs and others, we should always look for the best in their subject. Ensuring diversity in each of these roles is a deliberate process -- if there is not much diversity in the first set of names, it needs to be revised to include people from minoritized groups. If these groups are not showing up in the A-list, it is seldom because of their technical acumen -- it is because we, as organizers, are not used to looking in unfamiliar places to bring in the talent.  Many of our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyer_notice_2021, sarabipour_evaluating_2020}. Rather than deter us, this should thus encourage us to look further to find great people that are not routinely in the spotlight. We need to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard these people until there is ample representation across the diversity spectrum and dimensions.

% So, as long as we start with an inclusive and diverse organizing team (\textbf{Rule 2}), we can aim at finding the right people for various aspects of the conference organization. The regional and local communities are also good sources to tap into. For example, for useR! 2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential attendees, and sponsors.

%Disabled people often repeat "Nothing about us without us". This means the actual life experiences of people from URM are not replaceable with good intentions from people outside these groups. The same holds for other dimensions of diversity. No space will be truly inclusive and welcoming to URM without people from these groups bringing their experience to bear. This means people in privileged groups may need to step down from positions of power. Of course, no-one has the same experiences and experiences the same oppressions in life. %When building a diverse team, bear in mind that

@@
